New reduction techniques in commutative algebra driven by logical methods

### 2018-12-06 slides-padova2018.tex
Final changes

  Logic Seminar Padova \\
  December 5st, 2018
  \par
\end{frame}}

@@

  \jnote{2}{The displayed proof, which could have been taken from any
  standard textbook on commutative algebra, succeeds in this reduction
  by employing proof by contradition and minimal prime ideals. However, this way
  of reducing comes at a cost: It requires the Boolean Prime Ideal Theorem
  (for ensuring the existence of a prime ideal and for ensuring that stalks at
  minimal prime ideals are fields) and even the full axiom of choice (for
  ensuring the existence of a minimal prime ideal).

  We should hope that such a simple statement admits a more informative,
  explicit, computational proof: There should be an explicit method for
  transforming the given conditional equations expressing injectivity into the
  equation~$1 = 0$. And indeed there is: Beautiful constructive proofs can
  be found in Richman's note on
  \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{nontrivial
  uses of trivial rings} and in the
  \fixedhref{https://arxiv.org/abs/1605.04832}{recent textbook by Lombardi and
  Quitté} on constructive commutative algebra.

@@

  Unlike stalks or quotient rings, which are honest rings, the forcing
  model~$A^\sim$ is not a ring in the strict sense of the word: It doesn't
  have an underlying \emph{set} of elements, but instead an underlying
  \emph{sheaf} of elements. It is a \emph{ring object} in a certain category,
  the \emph{little Zariski topos} of~$A$. But as long as we restrict to intuitionistic
  reasoning, this difference is immaterial. Any intuitionistic theorem about
  rings applies to~$A^\sim$ as if~$A$ were a proper ring.

@@

  By~$A[f^{-1}]$, we mean the localization of~$A$ away from~$f$. This
  construction makes sense even if~$f$ is a zerodivisor, in which
  case~$A[f^{-1}]$ is the zero ring.}
\end{frame}

@@

  The definition of the semantics is reminiscient of Kripke and Beth models.
  Indeed, it is a fragment of the Kripke--Joyal semantics of the \emph{internal
  language of a topos}, and this general semantics encompasses Kripke and Beth
  models as special cases.}

@@

  \jnote{2-}{A free local ring~$A'$ over~$A$ is a local ring~$A'$
  together with a ring homomorphism~$A \to A'$ such that any ring
  homomorphim~$A \to R$ to a local ring uniquely factors over a local ring
  homomorphism~$A' \to R$. (A ring homomorphism is \emph{local} iff it reflects
  invertibility.)

@@

  \jnote{1}{Summarizing, we can associate to any reduced ring~$A$ the forcing
  model~$A^\sim$.
  \begin{itemize}
    \item It has the pleasant property that it is a field.
    \item Reasoning about it requires that we restrict ourselves to
    intuitionistic logic.
  \end{itemize}
  Depending on the application, this trade-off can not be useful at all, or be
  quite valuable. It certainly is so for proving Grothendieck's generic
  freeness lemma, simplifying a multi-page argument to a single and conceptual
  paragraph. (The previous slide only showed a fragment of the general
  statement of Grothendieck's generic freeness lemma. The technique outlined in
  these slides is amenable to the full version. Details are in Section~11.5 of
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
  notes}.)

>>>>

  Logic Seminar Padova \\
  December 5st, 2018
  \par

  \jnote{1}{We can associate to any reduced ring~$A$ a \emph{forcing
  model}~$A^\sim$.
  \begin{itemize}
    \item The forcing model has the pleasant property that it is a field.
    \item Reasoning about it requires that we restrict ourselves to
    intuitionistic logic.
  \end{itemize}
  Details are on the following slides.}
\end{frame}}

@@

  \jnote{2}{The displayed proof, which could have been taken from any
  standard textbook on commutative algebra, succeeds in this reduction
  by employing proof by contradiction and minimal prime ideals. However, this way
  of reducing comes at a cost: It requires the Boolean Prime Ideal Theorem
  (for ensuring the existence of a prime ideal and for ensuring that stalks at
  minimal prime ideals are fields) and even the full axiom of choice (for
  ensuring the existence of a minimal prime ideal).

  We should hope that such a simple statement admits a more informative,
  explicit, computational proof not employing transfinite methods: There should
  be an explicit method for
  transforming the given conditional equations expressing injectivity into the
  equation~$1 = 0$. And indeed there is: Beautiful constructive proofs can
  be found in Richman's note on
  \fixedhref{https://www.ams.org/journals/proc/1988-103-04/S0002-9939-1988-0954974-5/S0002-9939-1988-0954974-5.pdf}{nontrivial
  uses of trivial rings} and in the
  \fixedhref{https://arxiv.org/abs/1605.04832}{recent textbook by Lombardi and
  Quitté} on constructive commutative algebra.

@@

  Unlike stalks or quotient rings, which are honest rings, the forcing
  model~$A^\sim$ is not a ring in the strict sense of the word: It doesn't
  have an underlying \emph{set} of elements, but instead an underlying
  \emph{sheaf} of elements. It is a \emph{ring object} in a certain category,
  the \emph{little Zariski topos} of~$A$. But as long as we restrict to intuitionistic
  reasoning, this difference is immaterial. A metatheorem displayed on a later
  slide states that any intuitionistic theorem about
  rings applies to~$A^\sim$ just as if~$A$ were a proper, ordinary ring.

@@

  By~$A[f^{-1}]$, we mean the localization of~$A$ away from~$f$. This
  construction makes sense even if~$f$ is a zero divisor, in which
  case~$A[f^{-1}]$ is the zero ring.}
\end{frame}

@@

  The definition of the semantics is reminiscent of Kripke and Beth models.
  Indeed, it is a fragment of the Kripke--Joyal semantics of the \emph{internal
  language of a topos}, and this general semantics encompasses Kripke and Beth
  models as special cases.}

@@

  \jnote{2-}{A free local ring~$A'$ over~$A$ is a local ring~$A'$
  together with a ring homomorphism~$A \to A'$ such that any ring
  homomorphism~$A \to R$ to a local ring uniquely factors over a local ring
  homomorphism~$A' \to R$. (A ring homomorphism is \emph{local} iff it reflects
  invertibility.)

@@

  \jnote{1}{Summarizing, we can associate to any reduced ring~$A$ the forcing
  model~$A^\sim$.
  \begin{itemize}
    \item The forcing model has the pleasant property that it is a field.
    \item Reasoning about it requires that we restrict ourselves to
    intuitionistic logic.
  \end{itemize}
  Depending on the application, this trade-off can not be useful at all, or be
  quite valuable. It certainly is so for proving Grothendieck's generic
  freeness lemma, simplifying a multi-page argument to a single and conceptual
  paragraph. (The previous slide only showed a fragment of the general
  statement of Grothendieck's generic freeness lemma. The technique outlined in
  these slides is amenable to the full version. Details are in Section~11.5 of
  \fixedhref{https://rawgit.com/iblech/internal-methods/master/notes.pdf}{these
  notes}.)
